---
title: "Final"
author: "Ben Wollack"
date: "2026/06/09"
format:
  pdf:
    toc: true
editor: source
execute:
  message: false
  warning: false
---

# Set Up

The link to the GitHub repo can be found here\[https://github.com/benwollack/ENVS-193DS_spring-2026_final\].

```{r}
#| label: read-in

# read in packages
library(tidyverse)
library(here)
library(DHARMa)

# read in data
nest <- read_csv(here("data", "nest_data_final.csv"))
```

```{r}
#| label: height-bird
#| echo: FALSE

# visualizing tree height vs chance of bird use (case_control = 1)
ggplot(
  data = nest,
  aes(
    x = height_cm,
    y = case_control
  )
) +
  geom_jitter(width = 0, height = 0.2)
```

```{r}
#| label: ant-bird
#| echo: FALSE
#| output: FALSE

# visualizing ant species vs chance of bird use (case_control = 1)
# ignore NONE and TP, will be removed later
ggplot(
  data = nest,
  aes(
    x = ant_species,
    y = case_control
  )
) +
  geom_jitter(width = 0.1, height = 0.1)
```

```{r}
#| label: three-var
#| echo: FALSE
#| output: FALSE

# visualizing tree height and ant species vs chance of bird use
ggplot(
  data = nest,
  aes(
    x = ant_species,
    y = height_cm,
    color = case_control
  )) +
  geom_jitter(height = 0, 
              width = 0.2) +
  scale_color_gradient(
    low = "palevioletred",
    high = "blue"
  )
```

# Problem 1

## a.

In part 1, my coworker used a generalized linear model. I can tell because the results state a probability between two binary outcomes, the American Avocet using the habit or the American Avocet not using the habitat. This response variable is binary, which necessistates a generalized linear model to look at the relationship between that and the predictor of distance to the water.

In part 2, my coworker used a chi-squared test. I can tell because the results compare two categorical variables, habitat and bird species. Looking at the relationship between two categorical variables and seeing if one has an influence on the other requires using a chi-squared test.

## b.

For part 1: We found that distance to the water's edge influences how birds utilize microhabitats.

For part 2: We found a significant relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and the habitats they are found in (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

## c.

*Water depth* may influence the probability of microhabitat use. The microhabitats are all aquatic, and that water depth may influence things like the ability for birds to wade and the availabiliy of various food sources. Water depth is a continuous variable which can be measured in a unit such as cm or meters.

For the sake of data collection, it would be helpful to specify something specific and easily measurable like maximum water depth or average water depth after taking multiple measurements of the microhabitat.

# Problem 2

## a.

```{r}
#| label: data-cleaning

# create new object nest_clean from nest
nest_clean <- nest |> 
  # only keep the columns for ant species, tree height and nest presence (case_control)
  select(ant_species, height_cm, case_control) |> 
  # change species to full Latin names
  mutate(ant_species = case_when(
    ant_species == "RRB" ~ "C. mimosae",
    ant_species == "BBR" ~ "C. nigriceps",
    ant_species == "AB" ~ "C. sjostedti"
  )) |> 
  # remove NA values to only keep species we're interested in
  drop_na() |> 
  # assign levels to each ant species for later analysis
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                   "C. sjostedti",
                                   "C. mimosae",
                                   "C. nigriceps"
                                   ))
# view 10 random rows from the cleaned data frame
slice_sample(
  nest_clean,
  n = 10
)
```

## b.

The 1s in the dataset represent trees that have bird nests in them, and the 0s represent trees that do not have bird nests in them (called "available" in the study).

## c.

### Tree Height

Tree height is a continuous variable \[ask Prof Bui for a free link to study\]

### Acacia Ant Species

Acacia Ant Species is a categorical variable with the levels C. sjostedti, C. mimosae and C. nigriceps, which represent different species of Acacia Ants.

## d.

*+* indicates no interaction, *X* indicates an interaction

| Model number | Height | Ant Species |      Model Description       |
|--------------|:------:|:-----------:|:----------------------------:|
| 0            |   \+   |     \+      |  Both terms, no interaction  |
| 1            |   X    |      X      | Both terms, interaction term |

## e.

```{r}
#| label: model-run
#| output: FALSE

# model 0: glm with no interaction term
model0 <- glm(
  # formula for all predictors, no interaction
  case_control ~ height_cm + ant_species,
  # use cleaned data frame for nest data
  data = nest_clean,
  # set family to binomial due to the binomial response variable
  family = "binomial"
)

# model 1: glm with ant species as an interaction term with height
model1 <- glm(
  # asterisk to indicate an interaction
  case_control ~ height_cm * ant_species,
  # use cleaned data
  data = nest_clean,
  # set family to binomial due to the binomial response variable
  family = "binomial"
)
```

## f.

```{r}
#| label: AICc

# put both models into calculating AICc
AICc(
  model0,
  model1
) |> 
  # sorts the AICc values so the lowest (best model) appears first
  arrange(AICc)
```

The AICc for the model with an interaction term is 238.12 while the one without an interaction term is 258.89. The model with an interaction term is significantly higher (that is, it exceeds the other model by more than 2), so this is the best model for the system.

The best model as determined by Akaike’s Information Criterion (AIC) was the model where Acacia Ant species and tree height interact to determine the probability of a bird nest being present in the tree. The effect of tree height on the probability of a nest varies based on the ant species present in the tree.

## g.

```{r}
#| label: diagnostics

# display diagnostics for the glm with the interaction term
plot(
  simulateResiduals(model1)
  )
```

The QQ plot residuals roughly follow the reference line for a uniform distribution. The residuals vs predicted plot appears random with no noticeable patterns. From this, both diagnostic plots show that the glm assumptions are met.

## h.

## i.

## j.

## k.

# Problem 3

## a.

## b.

## c.

## d.

## e.

## f.

## g.

Ask Prof Bui
- Paper access
- Clarification on how many/which models to run (just 2, or do I need the null and single variable ones too?)
- Want to make sure my justification for choosing the interaction term is correct here
